dispRity demo

### 2015-10-06 doc/dispRity-tutorial.Rnw
Minor updates on tutorial

This a quick demo to go through the beta version of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).
Many parts are still missing (see section \ref{whatsleft}) but this version should be running all right and give an idea of what's implemented in the package.

@@

<<data, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Loading demo and the package data
library(dispRity)

## An ordinated matrix with the tips only
data(BeckLee_mat50)
dim(BeckLee_mat50)
head(BeckLee_mat50[,1:5])

## An ordinated matrix with the tips and the nodes (named nX)
data(BeckLee_mat99)
dim(BeckLee_mat99)
head(BeckLee_mat99[,1:5], 2)
tail(BeckLee_mat99[,1:5], 2)

## A list of first and last occurrences data for some fossils
data(BeckLee_ages)
head(BeckLee_ages)

## And a phylogeny:
data(BeckLee_tree)
plot(BeckLee_tree) ; nodelabels(cex=0.8)
@

@@

<<Setup the data>>=
## My ordinated matrix with tips only
MyData_mat50 <- BeckLee_mat50

## My ordinated matrix with tips and nodes
MyData_mat99 <- BeckLee_mat99

## My first and last occurence data
MyData_ages <- BeckLee_ages

## My phylogenetic tree
MyData_tree <- BeckLee_tree
@

@@

<<Quick go, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
## Splitting the data
sliced_data <- time.series(MyData_mat99, MyData_tree, method = "continuous",
    model = "acctran", time = 5, FADLAD = MyData_ages)

## Bootstrapping the data
bootstrapped_data <- boot.matrix(sliced_data, 100)

## Calculating disparity
sum_of_ranges <- dispRity(bootstrapped_data, metric = c(sum, ranges))

## Summarising the results
summary(sum_of_ranges)
plot(sum_of_ranges, type = "continuous")
@

@@

<<dispRity class2>>=
## Which class is the sum_of_ranges object?
class(sum_of_ranges)

## We can summarise it using the S3 method print.dispRity
sum_of_ranges
## Some attentive ape package users will note the similarities with a phylo object...
@

@@

<<cust.series2>>=
## We want to separate the species around the node 71 (see phylogeny above).
## All the descendant of these node are crown and all the ancestors are stem
crown <- extract.clade(MyData_tree, node = 71)$tip.label
stem <- drop.tip(MyData_tree, tip = crown)$tip.label

## We then have to feed this information in a data frame with one column
factors <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
    ncol = 1, dimnames = list(c(crown, stem))))

## We then can use the customised series function
crown_stem <- cust.series(MyData_mat50, factors)

## This created a dispRity object containing two series: crown and stem
class(crown_stem)
crown_stem

## This object contains three elements
names(crown_stem)

## With "data" being the list sub-matrices
str(crown_stem$data)

## "taxa" being the list of taxa in the original ordinated matrix
str(crown_stem$taxa)

## and finally, "series" containing information on the series type (custom) and names
## (crown and stem)
crown_stem$series
@

@@

<<discrete.series>>=
## Let's generate three time bins containing the data present every 40 Ma
time_bins <- time.series(data = MyData_mat50, tree = MyData_tree,
    method = "discrete", time = c(120, 80, 40, 0))

## Note that the function provides a warning saying that tips where single
## points in time (no FAD/LAD information). We can fix that by adding the
## age data for the taxa that have some longer occurrence spans.
time_bins <- time.series(data = MyData_mat50, tree = MyData_tree,
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = MyData_ages)

## To entirely avoid the warning we could collect the occurrence span data
## for all the taxa but that's not necessary. The function automatically
## assumes no occurrence span time (i.e. single points in time) for all taxa
## by default.

time_bins
@

@@

time_slices

## Note that in the same way as for the discrete method, we can also
## automatically generate the slices
time.series(data = MyData_mat99, tree = MyData_tree, method = "continuous",
    model = "gradual", time = 4)
## But that gives the same silly results (slicing every 44.50368 Ma)
@

@@

<<boot.matrix3>>=
## Let's do a full rarefaction on the crown/stem data for later
crown_stemBS <- boot.matrix(crown_stem, rarefaction = TRUE)
## And just bootstrapping for the time binning/slicing
time_binsBS <- boot.matrix(time_bins)
time_slicesBS <- boot.matrix(time_slices)

## And just to remind us what's in each objects
crown_stemBS
time_binsBS
time_slicesBS
@

@@

Let's start with plotting the difference in disparity between the crown and the stem mammals.
For this we are going to simply specify the type of plots using the \texttt{type} option that allows to choose between the \texttt{"continuous"} or \texttt{"discrete"} method.
In our case, we are interested in looking at the results in a discrete way.

@@

<<plot1, fig.width=15, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 15, height = 5) ; par(mfrow = (c(1,3)), bty = "n")
## Plotting the score for both groups
plot(disp_crown_stemBS, type = "discrete" , ylim = c(25,40))
## Same one with the number of taxa for each data set
plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, ylim = c(25,40))
## Same one with rarefied data (note how that affects the diversity data as well!)
plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, rarefaction = 20,
    ylim = c(25,40))
@

@@

<<plot2, fig.width=5, fig.height=8, out.width='0.5\\linewidth'>>=
## Graphical options
par(bty = "n")
## The rarefaction curves
plot(disp_crown_stemBS, type = "discrete", diversity = TRUE, rarefaction = "plot")
## Note that the function automatically draws the curves for each series and splits
## the plot screen accordingly.
@

@@

<<plot3, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
## The disparity-through-time data in a time-binned way (with lines rather than
## boxes)
plot(disp_time_binsBS, type = "discrete", discrete_type = "line")
## The disparity-through-time data in a time-sliced ways
plot(disp_time_slicesBS, type = "continuous")

@@

<<plot4, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
## Graphical options
par(bty = "n")
## First, plotting the continuous data
plot(disp_time_slicesBS, type = "continuous", ylim=c(0.5,2))
## Second, adding the discrete data
par(new = TRUE)
plot(disp_time_binsBS, type = "discrete", discrete_type = "line", ylim=c(0.5,2),
    xlab="", ylab="")
@

>>>>

This is a quick demo to go through the beta version of the \dispRity package (v.\Sexpr{packageVersion("dispRity")}).
Many parts are still missing (see section \ref{whatsleft}) but this version should be running all right and give an idea of what's implemented in the package.

@@

<<data, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Loading demo and the package data
library(dispRity)
## An ordinated matrix with the tips only
data(BeckLee_mat50)
dim(BeckLee_mat50)
head(BeckLee_mat50[,1:5])
## An ordinated matrix with the tips and the nodes (named nX)
data(BeckLee_mat99)
dim(BeckLee_mat99)
head(BeckLee_mat99[,1:5], 2)
tail(BeckLee_mat99[,1:5], 2)
## A list of first and last occurrences data for some fossils
data(BeckLee_ages)
head(BeckLee_ages)
## And a phylogeny:
data(BeckLee_tree)
plot(BeckLee_tree) ; nodelabels(cex=0.8)
@

@@

<<Setup the data>>=
## My ordinated matrix with tips only
MyData_mat50 <- BeckLee_mat50
## My ordinated matrix with tips and nodes
MyData_mat99 <- BeckLee_mat99
## My first and last occurence data
MyData_ages <- BeckLee_ages
## My phylogenetic tree
MyData_tree <- BeckLee_tree
@

@@

<<Quick go, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
## Splitting the data
sliced_data <- time.series(MyData_mat99, MyData_tree, method = "continuous",
    model = "acctran", time = 9, FADLAD = MyData_ages)
## Bootstrapping the data
bootstrapped_data <- boot.matrix(sliced_data, 100)
## Calculating disparity
sum_of_ranges <- dispRity(bootstrapped_data, metric = c(sum, ranges))
## Summarising the results
summary(sum_of_ranges)
plot(sum_of_ranges)
@

@@

<<dispRity class2>>=
## Which class is the sum_of_ranges object?
class(sum_of_ranges)
## We can summarise it using the S3 method print.dispRity
sum_of_ranges
## Some attentive ape package users will note the similarities with a phylo object...
@

@@

<<cust.series2>>=
## We want to separate the species around the node 71 (see phylogeny above).
## All the descendant of these node are crown and all the ancestors are stem
crown <- extract.clade(MyData_tree, node = 71)$tip.label
stem <- drop.tip(MyData_tree, tip = crown)$tip.label
## We then have to feed this information in a data frame with one column
factors <- as.data.frame(
    matrix(data = c(rep("crown", length(crown)), rep("stem", length(stem))),
    ncol = 1, dimnames = list(c(crown, stem))))
## We then can use the customised series function
crown_stem <- cust.series(MyData_mat50, factors)
## This created a dispRity object containing two series: crown and stem
class(crown_stem)
crown_stem
## This object contains three elements
names(crown_stem)
## With "data" being the list sub-matrices
str(crown_stem$data)
## "taxa" being the list of taxa in the original ordinated matrix
str(crown_stem$taxa)
## and finally, "series" containing information on the series type (custom) and names
## (crown and stem)
crown_stem$series
@

@@

<<discrete.series>>=
## Let's generate three time bins containing the data present every 40 Ma
time_bins <- time.series(data = MyData_mat50, tree = MyData_tree,
    method = "discrete", time = c(120, 80, 40, 0))
## Note that the function provides a warning saying that tips where single
## points in time (no FAD/LAD information). We can fix that by adding the
## age data for the taxa that have some longer occurrence spans.
time_bins <- time.series(data = MyData_mat50, tree = MyData_tree,
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = MyData_ages)
## To entirely avoid the warning we could collect the occurrence span data
## for all the taxa but that's not necessary. The function automatically
## assumes no occurrence span time (i.e. single points in time) for all taxa
## by default.
time_bins
@

@@

time_slices
## Note that in the same way as for the discrete method, we can also
## automatically generate the slices
time.series(data = MyData_mat99, tree = MyData_tree, method = "continuous",
    model = "gradual", time = 4)
## But that gives the same silly results (slicing every 44.50368 Ma)
@

@@

<<boot.matrix3>>=
## Let's do a full rarefaction on the crown/stem data for later
crown_stemBS <- boot.matrix(crown_stem, rarefaction = TRUE)
## And just bootstrapping for the time binning/slicing
time_binsBS <- boot.matrix(time_bins)
time_slicesBS <- boot.matrix(time_slices)
## And just to remind us what's in each objects
crown_stemBS
time_binsBS
time_slicesBS
@

@@

Let's start with plotting the difference in disparity between the crown and the stem mammals.
For this we are going to simply specify the type of plots using the \texttt{type} option that allows to choose between the \texttt{"continuous"} or \texttt{"discrete"} method.
In our case, we are interested in looking at the results in a discrete way.
Note that the plot \texttt{type} can be left missing.
The function will then use \texttt{"discrete"} as default or \texttt{"continuous"} if the input data came from the \texttt{time.series} function with \texttt{method = "continuous"}.

@@

<<plot1, fig.width=15, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 15, height = 5) ; par(mfrow = (c(1,3)), bty = "n")
## Plotting the score for both groups
plot(disp_crown_stemBS, ylim = c(25,40))
## Same one with the number of taxa for each data set
plot(disp_crown_stemBS, diversity = TRUE, ylim = c(25,40))
## Same one with rarefied data (note how that affects the diversity data as well!)
plot(disp_crown_stemBS, diversity = TRUE, rarefaction = 20, ylim = c(25,40))
@

@@

<<plot2, fig.width=5, fig.height=8, out.width='0.5\\linewidth'>>=
## Graphical options
par(bty = "n")
## The rarefaction curves
plot(disp_crown_stemBS, diversity = TRUE, rarefaction = "plot")
## Note that the function automatically draws the curves for each series and splits
## the plot screen accordingly.
@

@@

<<plot3, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
## The disparity-through-time data in a time-binned way (with lines rather than
## boxes)
plot(disp_time_binsBS, discrete_type = "line")
## The disparity-through-time data in a time-sliced ways
plot(disp_time_slicesBS)

@@

<<plot4, fig.width=5, fig.height=5, out.width='.5\\linewidth'>>=
## Graphical options
par(bty = "n")
## First, plotting the continuous data
plot(disp_time_slicesBS, ylim=c(0.5,2))
## Second, adding the discrete data
par(new = TRUE)
plot(disp_time_binsBS, discrete_type = "line", ylim=c(0.5,2), xlab="", ylab="")
@
